In-silico molecular enrichment and clearance of the human intracranial space

### 2025-06-17 paper/main.tex
change R_red to to R_c^red etc
revision I by Marius III

% Variations with sleep, exercise
Lifestyle factors such as sleep~\cite{xie2013sleep, miao2024brain,
  bojarskaite2023sleep, eide2021sleep, vinje2023human,
  hauglund2025norepinephrine, larsen2024sleep},
exercise~\cite{holstein2018voluntary, yildiz2022immediate}, and
alcohol intake~\cite{lundgaard2018beneficial} affect molecular
enrichment and clearance of the brain. In particular, pial arteries
display higher-amplitude low-frequency vasomotion during NREM and
intermediate sleep states, while REM sleep is associated with reduced
PVS width~\cite{bojarskaite2023sleep}. Peristaltic pumping is more
effective under both of these configurations, resulting in higher
estimated net PVS flow velocities~\cite{gjerde2023directional}. Even
higher PVS velocities would amplify the PVS flow effects simulated
here, with earlier arrival times along the PVSs, clear enrichment in
adjacent spaces, and accelerated molecular enrichment
overall. Moreover, sleep is linked with low-frequency oscillations in
human CSF flow~\cite{fultz2019coupled}. As we have shown here, the
level of dispersion in the CSF spaces strongly shapes molecular
enrichment patterns. While we, at baseline, account for the dispersion
induced by cardiac and respiratory pulsatility, the high dispersion
scenarios considered can be viewed as representative of additional
pulsatility induced by e.g.~neural waves~\cite{fultz2019coupled,
  williams2023neural} or other respiration patterns depending on
activity level. More broadly, we consider this as a key strength of
the current in-silico framework: allowing for integrating and
exploring the mechanistic macroscale implications of changes in
pulsatility at smaller spatial or temporal scales.
This approach of modeling pulsatility through an effective dispersion factor is a simplification that does not capture the full complexity of transient fluid dynamics, for instance caused by non-uniform wall motion and ependymal cilia within the ventricular system~\cite{siyahhan2014flow}.
However, it is further supported by recent clinical observations. High-resolution MRI measurements by Hirschler et al.~\cite{hirschler2024region} revealed that CSF mobility in regions with high pulsatility is approximately tenfold greater than the self-diffusion of water, an enhancement factor consistent with the values computed in our model.

@@

    2 & reduced CSF pulsatility & $\bm u_{\rm CSF}$ & $\hat{u}_a^p + \hat{u}_a^c$ & $\hat{u}_v^p$  & \blue{$ R_{\rm red}$} & $\xi_{\rm EF}$ & $2 R_1$  \\ 
    3 & increased CSF pulsatility & $\bm u_{\rm CSF}$ & $\hat{u}_a^p + \hat{u}_a^c$ & $\hat{u}_v^p$  & \blue{$R_{\rm inc}$} & $\xi_{\rm EF}$ & $2 R_1$  \\ 

@@

    variations $R_{\rm red}$ and $R_{\rm inc}$ corresponding to

@@

We give a concise summary of the numerical approach here (see
Supplementary methods for a comprehensive description). To
solve~\eqref{eq:multi_transport} numerically, we use the interior
penalty discontinuous Galerkin (DG) method with weighted averages and
upwinding for the convection term in the 3D
domains~\cite{ern2009discontinuous} -- to ensure numerical stability
and minimize the artificial presence of negative concentration
values. For the 1D domain, we use continuous finite element spaces
defined over each of the networks and stabilized with numerical
diffusion when $\hat{u}$ is non--zero. Note that the 3D--1D coupling
in~\eqref{eq:multi_transport} leads to a 3D solution $c$ with low
regularity properties. Thus, one can only expect lower-order error
convergence near $\Lambda_a$ and $\Lambda_v$, see the references
\cite{masri2024discontinuous} and \cite{laurino2019derivation} for the
numerical analysis of DG and continuous Galerkin schemes,
respectively. However, away from these 1D networks, almost optimal
approximation is
expected~\cite{masri2023discontinuous,koppl2016local}. To numerically
solve the Stokes equations to compute the CSF flow velocity and
pressure in the ventricular system and SAS, we use finite element
spaces that preserve the incompressibility condition on the discrete
level~\cite{hong2016robust}. This requirement is important for
numerical stability when subsequently
solving~\eqref{eq:multi_transport}\cite{cesmelioglu2022compatible}.
The spaces we used for the velocity are continuous along the normal
direction; continuity of the tangential component is enforced via
interior penalty approaches~\cite{hong2016robust}. Optimal
approximation properties are expected for this scheme. These numerical
methods were implemented using the FEniCS finite element
framework~\cite{alnaes2015fenics}, and the 3D-1D coupling is handled
by the extension
FEniCS\textsubscript{ii}~\cite{kuchta2020assembly}. The correctness
and accuracy of the numerical solutions were verified by a series of
numerical verification experiments (see Supplementary information S1.5).

>>>>

% Variations with sleep, exercise
Lifestyle factors such as sleep~\cite{xie2013sleep, miao2024brain,
  bojarskaite2023sleep, eide2021sleep, vinje2023human,
  hauglund2025norepinephrine, larsen2024sleep},
exercise~\cite{holstein2018voluntary, yildiz2022immediate}, and
alcohol intake~\cite{lundgaard2018beneficial} affect molecular
enrichment and clearance of the brain. In particular, pial arteries
display higher-amplitude low-frequency vasomotion during NREM and
intermediate sleep states, while REM sleep is associated with reduced
PVS width~\cite{bojarskaite2023sleep}. Peristaltic pumping is more
effective under both of these configurations, resulting in higher
estimated net PVS flow velocities~\cite{gjerde2023directional}. Even
higher PVS velocities would amplify the PVS flow effects simulated
here, with earlier arrival times along the PVSs, clear enrichment in
adjacent spaces, and accelerated molecular enrichment
overall. Moreover, sleep is linked with low-frequency oscillations in
human CSF flow~\cite{fultz2019coupled}. As we have shown here, the
level of dispersion in the CSF spaces strongly shapes molecular
enrichment patterns. While we, at baseline, account for the dispersion
induced by cardiac and respiratory pulsatility, the high dispersion
scenarios considered can be viewed as representative of additional
pulsatility induced by e.g.~neural waves~\cite{fultz2019coupled,
  williams2023neural} or other respiration patterns depending on
activity level. More broadly, we consider this as a key strength of
the current in-silico framework: allowing for integrating and
exploring the mechanistic macroscale implications of changes in
pulsatility at smaller spatial or temporal scales.
This approach of modeling pulsatility through an effective dispersion 
factor is a simplification that does not capture the full complexity
of transient fluid dynamics, such as non-uniform wall motion and ependymal
cilia within the ventricular system~\cite{siyahhan2014flow}, dynamic CSF
flow at the craniocervical junction, or the effect of microanatomical
features such as subarachnoid trabeculae on flow resitance and dispersion
in the SAS. However, our approach is supported by recent clinical 
observations. High-resolution MRI measurements by Hirschler et
al.~\cite{hirschler2024region} revealed that CSF mobility in regions
with high pulsatility is approximately tenfold greater than the 
self-diffusion of water, an enhancement factor consistent with 
the values computed in our model.

@@

    2 & reduced CSF pulsatility & $\bm u_{\rm CSF}$ & $\hat{u}_a^p + \hat{u}_a^c$ & $\hat{u}_v^p$  & \blue{$ R_c^{\rm red} + R_r^{\rm red}$} & $\xi_{\rm EF}$ & $2 R_1$  \\ 
    3 & increased CSF pulsatility & $\bm u_{\rm CSF}$ & $\hat{u}_a^p + \hat{u}_a^c$ & $\hat{u}_v^p$  & \blue{$ R_c^{\rm inc} + R_r^{\rm inc}$} & $\xi_{\rm EF}$ & $2 R_1$  \\ 

@@

    variations $R_{r/c}^{\rm red}$ and $R_{r/c}^{\rm inc}$ corresponding to

@@

We give a concise summary of the numerical approach here (see
Supplementary methods for a comprehensive description). To
solve~\eqref{eq:multi_transport} numerically, we use the interior
penalty discontinuous Galerkin (DG) method with weighted averages and
upwinding for the convection term in the 3D
domains~\cite{ern2009discontinuous} -- to ensure numerical stability
and minimize the artificial presence of negative concentration
values. For the 1D domain, we use continuous finite element spaces
defined over each of the networks and stabilized with numerical
diffusion when $\hat{u}$ is non--zero. Note that the 3D--1D coupling
in~\eqref{eq:multi_transport} leads to a 3D solution $c$ with low
regularity properties. Thus, one can only expect lower-order error
convergence near $\Lambda_a$ and $\Lambda_v$, see the references
\cite{masri2024discontinuous} and \cite{laurino2019derivation} for the
numerical analysis of DG and continuous Galerkin schemes,
respectively. However, away from these 1D networks, almost optimal
approximation is
expected~\cite{masri2023discontinuous,koppl2016local}. To numerically
solve the Stokes equations to compute the CSF flow velocity and
pressure in the ventricular system and SAS, we use finite element
spaces that preserve the incompressibility condition on the discrete
level~\cite{hong2016robust}. This requirement is important for
numerical stability when subsequently
solving~\eqref{eq:multi_transport}\cite{cesmelioglu2022compatible}.
The spaces we used for the velocity are continuous along the normal
direction; continuity of the tangential component is enforced via
interior penalty approaches~\cite{hong2016robust}. Optimal
approximation properties are expected for this scheme. These numerical
methods were implemented using the FEniCS finite element
framework~\cite{alnaes2015fenics}, and the 3D-1D coupling is handled
by the extension FEniCS\textsubscript{ii}~\cite{kuchta2020assembly}.
The correctness and accuracy of the numerical solutions were verified by a series of
numerical verification experiments (see Supplementary information S1.5).
The majority of the simulation time was spend on the Stokes flow and transport model simulations. Each Stokes flow simulation utilizes a hybrid parallel approach with 16 MPI processes and 4 threads each, requiring approximately 300 GB of RAM and completing in 23 minutes. The transport model is limited to serial execution by FEniCS\textsubscript{ii}, but employes 16 threads for the linear system factorization, taking 22 minutes and consuming about 50 GB of memory on a AMD EPYC 9684X machine.
